Verifying Concurrent Search Structure Templates for C Programs using VST

### 2023-04-09 concurrency/CONCUR 2023/paper.tex
WIP on verified

Figure \ref{traverse_lock} presents two versions of the \lstinline{traverse} function: one written in C (Figure \ref{traverse_lock_a}) and the other in an ML-like language (Figure \ref{traverse_lock_b}). The algorithm assumes that certain helper functions must be provided by the implementation that satisfy certain criteria. One such function, namely \lstinline{findNext}, is utilized by the algorithm to determine the next node $n'$ to be visited based on the current node $n$ and the key $k$ belonging to the set of key values.

@@

\begin{figure}[ht]
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		\label{traverse_lock_a}
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in ML-like language \cite{krishna2021automated}}
		\label{traverse_lock_b}	
	\end{subfigure}
	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms}
	\label{traverse_lock}
\end{figure}

To implement the template methodology for concurrent data structure implementations in C, we have implemented the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}. The \lstinline{pn} struct includes two fields, namely \lstinline{p} and \lstinline{n}, which respectively represent the current and next node. Additionally, we require a target key \lstinline{k} and a dummy node value \lstinline{value} (referred to as 'dummy' because it is not actually used in \lstinline{traverse}). We include this dummy node to maintain consistency with other implementations. Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language, as it relies on a helper function called \lstinline{findnext}. The purpose of \lstinline{findnext} is to determine the next node to be visited during the traversal. In our implementation, the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}. Alternatively, it terminates if \lstinline{findnext} fails to find a next node that includes the key value $k$ in the set of key values. The \lstinline{traverse} function returns a Boolean value indicating whether it has successfully found the next node during traversal. If it fails to do so, it returns false, which other functions can use to make decisions.

For instance, in a binary search tree, a node with a value of 20 would have a left child with a set of key values consisting of values less than 20 and a right child with one greater than 20. If we want to insert a new node with the target key $k$ into the tree, we can use the \lstinline{traverse} function to traverse the tree. When \lstinline{traverse} fails to find the next node (either to the left or right), it implies that it has reached a node with a target key $k$ that equals the node's value. At this point, we only have to modify the value of that node to \lstinline{value}.

@@

Figure \ref{traverse_giveup_b} depicts the give-up template algorithm, which is presented in an ML-like language and described in \cite{krishna2021automated}. The give-up template algorithm uses a helper function called \lstinline{inRange}, in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section. The purpose of the \lstinline{inRange} function is to determine whether the key value $k$ falls within the range of node $n$. If the key value is outside the node's range, the search is terminated, and the thread is restarted.

@@

%% C implementation 
To implement the template algorithm, it is crucial to keep track of the lower and upper bounds of the keys present in the subtree rooted at the current node. Therefore, we have incorporated two new fields, \lstinline{min} and \lstinline{max}, into each node of type \lstinline{tree_t} in our C implementation. These fields store the lower and upper bounds of the keys, respectively. When a thread arrives at a node in search of the key $k$, it checks whether $k$ falls within the range of that node. If $k$ falls within the range, the thread continues by calling \lstinline{findnext} to find the new node. If $k$ does not fall within the range, the thread releases the lock and returns to the root node by assigning the pointer of the root to the pointer of the current node (lines 21-22 in Figure \ref{traverse_giveup_a}).

@@

For many concurrent data structures, the ideal correctness condition requires that the data structure behaves identically to a sequential implementation, even when accessed concurrently by multiple threads. The concurrent behavior of search structures is specified using \emph{atomic triples}~\cite{tada}. These take the form of $\forall a.\ \left\langle P_l\ |\ P_p(a) \right\rangle\ c\ \left\langle Q_l\ |\ Q_p(a)\right\rangle$, where $P_l$ and $Q_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $P_p$ and $Q_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure. This triple suggests that if $P_l$ holds true before a call and $P_p$ is true for some value of $a$ in a shared state (e.g. a global invariant), then $P_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $c$, at which point $Q_p$ will become true atomically for the same value $a$ (and $Q_l$ will be true after $c$ ends). For instance, the specification
$\forall s.\ \left\langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\right\rangle\ \texttt{push}(v)\ \left\langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\right\rangle$
expresses the fact that the push operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from particular stack $s$ to $v::s$ at some point during its execution.

@@

$$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn)} \ast \lockinv \texttt{(pn->p->lock, R'}) \ast \texttt{R(pn->p)} \ | \ \treerep\ \texttt{t} \right\rangle $$
$$\texttt{traverse(pn, k)}$$
$$\left\langle v.\ \nodeboxrep \texttt{(pn)} \ast \lockinv \texttt{(pn->n->lock, R')} \ast \texttt{R(pn->p)} \ast (\texttt{k} \in \texttt{range})
\ |\ \treerep\ \texttt{t} \right\rangle$$

We need instantiate $\nodeboxrep$ (the resources associated with each client thread) and $\treerep$ (the abstract representation of the data structure), and show how they interact to access and update the state of the data structure for each call. In a lock-based implementation, $\nodeboxrep$ serves as a pointer to the root node's lock, and acquiring this lock allows the thread to access more elements within the data structure. As previously mentioned, the \lstinline{pn} struct comprises two nodes of type \lstinline{tree_t}: the \lstinline{p} field, which denotes the current node, and the \lstinline{n} field, which represents the next node. The resources $\texttt{R}$, parameterized by a node pointer, describes the concrete information about a node. The $\lockinv$ asserts that there exists a lock in memory with a given invariant.

>>>>

Figure \ref{traverse_lock} presents two versions of the \lstinline{traverse} function: one written in C (Figure \ref{traverse_lock_a}) and the other in an ML-like language (Figure \ref{traverse_lock_b}). The algorithm assumes that certain helper functions must be provided by the implementation that satisfy certain criteria. One such function, namely \lstinline{findNext}, is utilized by the algorithm to determine the next node $\texttt{n'}$ to be visited based on the current node $\texttt{n}$ and the key $\texttt{k}$ belonging to the set of key values.

@@

\begin{figure}[ht]
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		\label{traverse_lock_a}
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in ML-like language \cite{krishna2019compositional}}
		\label{traverse_lock_b}	
	\end{subfigure}
	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms}
	\label{traverse_lock}
\end{figure}

To implement the template methodology for concurrent data structure implementations in C, we have implemented the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}. The \lstinline{pn} struct includes two fields, namely \lstinline{p} and \lstinline{n}, which respectively represent the current and next node. Additionally, we require a target key \lstinline{k} and a dummy node value \lstinline{value} (referred to as 'dummy' because it is not actually used in \lstinline{traverse}). We include this dummy node to maintain consistency with other implementations. Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language, as it relies on a helper function called \lstinline{findnext}. The purpose of \lstinline{findnext} is to determine the next node to be visited during the traversal. In our implementation, the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}. Alternatively, it terminates if \lstinline{findnext} fails to find a next node that includes the key value $\texttt{k}$ in the set of key values. The \lstinline{traverse} function returns a Boolean value indicating whether it has successfully found the next node during traversal. If it fails to do so, it returns false, which other functions can use to make decisions.

For instance, in a binary search tree, a node with a value of 20 would have a left child with a set of key values consisting of values less than 20 and a right child with one greater than 20. If we want to insert a new node with the target key $\texttt{k}$ into the tree, we can use the \lstinline{traverse} function to traverse the tree. When \lstinline{traverse} fails to find the next node (either to the left or right), it implies that it has reached a node with a target key $\texttt{k}$ that equals the node's value. At this point, we only have to modify the value of that node to \lstinline{value}.

@@

Figure \ref{traverse_giveup_b} depicts the give-up template algorithm, which is presented in an ML-like language and described in \cite{krishna2019compositional}. The give-up template algorithm uses a helper function called \lstinline{inRange}, in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section. The purpose of the \lstinline{inRange} function is to determine whether the key value $k$ falls within the range of node $n$. If the key value is outside the node's range, the search is terminated, and the thread is restarted.

@@

%% C implementation 
To implement the template algorithm, it is crucial to keep track of the lower and upper bounds of the keys present in the subtree rooted at the current node. Therefore, we have incorporated two new fields, \lstinline{min} and \lstinline{max}, into each node of type \lstinline{tree_t} in our C implementation. These fields store the lower and upper bounds of the keys, respectively. When a thread arrives at a node in search of the key $\texttt{k}$, it checks whether $\texttt{k}$ falls within the range of that node. If $\texttt{k}$ falls within the range, the thread continues by calling \lstinline{findnext} to find the new node. If $\texttt{k}$ does not fall within the range, the thread releases the lock and returns to the root node by assigning the pointer of the root to the pointer of the current node (lines 21-22 in Figure \ref{traverse_giveup_a}).

@@

For many concurrent data structures, the ideal correctness condition requires that the data structure behaves identically to a sequential implementation, even when accessed concurrently by multiple threads. The concurrent behavior of search structures is specified using \emph{atomic triples}~\cite{tada}. These take the form of $\forall a.\ \left\langle P_l\ |\ P_p(a) \right\rangle\ \texttt{c}\ \left\langle Q_l\ |\ Q_p(a)\right\rangle$, where $P_l$ and $Q_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $P_p$ and $Q_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure. This triple suggests that if $P_l$ holds true before a call and $P_p$ is true for some value of $a$ in a shared state (e.g. a global invariant), then $P_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $Q_p$ will become true atomically for the same value $a$ (and $Q_l$ will be true after $\texttt{c}$ ends). For instance, the specification
$\forall s.\ \left\langle \texttt{is\_stack}\ p\ |\ \texttt{stack}\ s\right\rangle\ \texttt{push}(v)\ \left\langle \texttt{is\_stack}\ p\ |\ \texttt{stack}\ (v::s)\right\rangle$
expresses the fact that the push operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from particular stack $s$ to $v::s$ at some point during its execution.

@@

$$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn)} \ast \lockinv \texttt{(pn->p->lock,R'}) \ast \texttt{R(pn->p)} \ | \ \treerep\ \texttt{t} \right\rangle $$
$$\texttt{traverse(pn, k)}$$
$$\left\langle \texttt{n'}.\ \nodeboxrep \texttt{(pn)} \ast \lockinv \texttt{(pn->n'->lock,R')} \ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range})
\ |\ \treerep\ \texttt{t} \right\rangle$$

We need to instantiate $\nodeboxrep$ (the resources associated with each client thread) and $\treerep$ (the abstract representation of the data structure), and show how they interact to access and update the state of the data structure for each call. In a lock-based implementation, $\nodeboxrep$ serves as a pointer to the root node's lock, and acquiring this lock allows the thread to access more elements within the data structure. As previously mentioned, the \lstinline{pn} struct comprises two nodes of type \lstinline{tree_t}: the \lstinline{p} field, which denotes the current node, and the \lstinline{n} field, which represents the next node. The resources $\texttt{R}$, parameterized by a node pointer, describes the concrete information about a node. The lock invariant $\lockinv$ predicate asserts that there exists a lock that protects the resources.
